survMarkerTwoPhase
=============================================

This R package computes non-parametric and semi-parametric estimates of accuracy measures for risk prediction markers from survival data under two phase study designs, namely the case-cohort (cch) and the nested case-control (ncc) study designs. 

The accuracy measures that can be estimated include the **AUC**, **TPR(c)**, **FPR(c)**, **PPV(c)**, and **NPV(c)** for for a specific timepoint and fixed marker cutoff value c. Standard errors, along with normal approximated confidence intervals are also calculated. *Note: standard errors for non-parametric estimates from a cch design are currently not available*.  

Below is a brief tutorial to get you started. For more information regarding estimation procedures see the references below. 


##Tutorial

### Download and install the package


```{r, eval=FALSE}
#download the package from github
if (!require("devtools")) install.packages("devtools")
devtools::install_github("mdbrown/survMarkerTwoPhase")
```

```{r, message = FALSE}
library(survMarkerTwoPhase)

#simulated data for illustration
data(SimData)  
head(SimData)
```



### Case-cohort design

For the case-cohort design we sample **n=150** from entire cohort, and include all participants with observed failures. 

```{r}
#generate a sub-cohort from SimData
set.seed(12321)
#create a sample index. 1 if sampled, 0 if not
N <- nrow(SimData)
sampleInd <- rep(0, N)

# sample all with observed failure time. (200 individuals)
sampleInd[SimData$status==1] <- 1

#sample 150 more observations from the entire data set without replacement
sampleInd[sample(1:N, 150)] <- 1

table(sampleInd) #total number of subcohort is 293 
```
To estimate accuracy measures for the case-cohort design, sample weights must be calculated. The sample weights *w = 1/Pr(Sampled from cohort)* for each observation included in the sub-cohort. 
```{r}
cohortData_cch <- SimData

#first calculate the Pr(Sampled from cohort) for each observation
sampleProb <- numeric(500)

#all non-censored observations were sampled, so their sample probability is 1
sampleProb[cohortData_cch$status==1] <- 1 
#all other individuals had a 150/N chance to be sampled 
sampleProb[cohortData_cch$status==0] <- 150/N

#the sample weights are 1/(probability of being sampled)
cohortData_cch$weights <- 1/sampleProb

#indicator of inclusion in the subcohort
cohortData_cch$subcohort = sampleInd  

#marker data is unavailable for those not in the subcohort
cohortData_cch$Y[sampleInd==0] = NA  

```

Here we estimate using non-parametric estimation methods using double inverse probability weighting.   

```{r}
#estimate accuracy measures using non-parametric estimates
#by setting estimation.method = "NP"
survMTP.cch(time =survTime, 
            event = status, 
            marker = Y,
            weights = weights,
            subcoh = subcohort,
            data = cohortData_cch, 
            estimation.method = "NP",  
            predict.time = 2, 
            marker.cutpoint = 0)

```

Now estimate measures  using semi-parametric estimation methods. These methods assume a Cox proportional hazards model.

```{r}
#estimate accuracy measures using semi-parametric estimates
survMTP.cch(time =survTime, 
            event = status, 
            marker = Y,
            weights = weights,
            subcoh = subcohort,
            data = cohortData_cch, 
            estimation.method = "SP",  
            predict.time = 2, 
            marker.cutpoint = 0)

```


For more information see `?survMTP.cch`. 

### Nested case-control
For a nested case-control sample, we need to input the full cohort data, with an indicator for inclusion in the subcohort, as well as a dataframe of risk sets, where each row is a case id followed by the corresponding id's that were matched to it. 

Generate a nested case-control subcohort using the function `ccwc` from the [`Epi`](http://cran.r-project.org/web/packages/Epi/index.html) package. 

```{r, results='hide', message=FALSE, warning=FALSE, tidy=TRUE}
require('Epi')

#2 matched controls for each case
nmatch = 2

cohortData_ncc <- SimData
cohortData_ncc$id = 1:dim(cohortData_ncc)[1]

subcohort_ncc <- ccwc(exit = survTime, fail= status, data = cohortData_ncc, controls = nmatch) # match 2 controls for each case. 

#indicator for inclusion in the subcohort
sampleInd = rep(0, nrow(cohortData_ncc)) #initialize all equal to zero
sampleInd[subcohort_ncc$Map] = 1

cohortData_ncc$subcohort = sampleInd  

table(sampleInd) #subcohort sample size of 352

#marker data is unavailable for those not in the subcohort
cohortData_ncc$Y[sampleInd==0] = NA  

#now we need to build the set matrix,
#which will be dimension (# of cases) x (nmatch + 1), so 200x3 here
#each row denotes a selected set, with the case id followed by the matched control ids
Sets <- matrix(nrow = sum(cohortData_ncc$status), ncol = nmatch +1)

for(i in subcohort_ncc$Set){
  Sets[i, ] <- unlist(subset(subcohort_ncc, Set==i, select=Map))
}

```

Now we are ready to calculate the measures using `survMTP.ncc`. Standard errors for both SP and NP estimates are calculated using perturbation. NP estimates use kernel smoothing while SP estimates assume a Cox proportional hazards model. 

```{r}
#Nonparametric estimates 
survMTP.ncc(time = survTime, 
         event = status, 
         marker=Y, 
         subcoh=subcohort, 
         id = id, 
         data = cohortData_ncc,
         sets = Sets, 
         estimation.method = "NP", 
         predict.time = 2 , 
         marker.cutpoint = 0)
```

```{r}
#Semiparametric estimates 
survMTP.ncc(time = survTime, 
         event = status, 
         marker=Y, 
         subcoh=subcohort, 
         id = id, 
         data = cohortData_ncc,
         sets = Sets, 
         estimation.method = "SP", 
         predict.time = 2 , 
         marker.cutpoint = 0)
```
 
See `?survMTP.ncc` for further help. 

### References

1. Liu D, Cai T, Zheng Y. Evaluating the predictive value of biomarkers with stratified case-cohort design. *Biometrics* 2012, 4: 1219-1227.

2. Pepe MS, Zheng Y, Jin Y. Evaluating the ROC performance of markers for future events. *Lifetime Data Analysis.* 2008, 14: 86-113.

3. Zheng Y, Cai T, Pepe MS, Levy, W. Time-dependent predictive values of prognostic biomarkers with failure time outcome. *JASA* 2008, 103: 362-368.

4. Cai T. and Zheng Y . Resampling Procedures for Making Inference under Nested Case-control Studies. *JASA* 2013 (in press). 

5. Cai T and Zheng Y/*, Evaluating Prognostic Accuracy of Biomarkers under nested case-control studies. *Biostatistics* 2012, 13,1, 89-100. 

(* equal contributor and corresponding author). 











