survMarkerTwoPhase
=============================================

This R package computes semi-parametric estimates of accuracy measures for risk prediction markers from survival data under two phase designs. For now, estimates can be obtained for the the case-cohort study design. Calculations for the nested case-control design will be implemented soon.

Once sample weights are obtained, the function `survMTP.estimate` estimates the $AUC$, $TPR(c)$, $FPR(c)$, $PPV(c)$, and $NPV(c)$ for for a specific timepoint and marker cutoff value c. Standard errors, and confidence intervals are also computed. 

For more information, see references below. 


### Download and install the package


```{r}
#download the package from github
if (!require("devtools")) install.packages("devtools")
devtools::install_github("survMarkerTwoPhase", "mdbrown")


library(survMarkerTwoPhase)

#simulated data for illustration
data(SimData)
head(SimData)
```



### Case-Cohort Design

Estimation using a case-cohort subcohort design is permitted. Sample weights must first be calculated based.

####Type I: 
 
Sample $n=150$ from entire cohort, and include all participants with observed failures. 

```{r}
#generate a sub-cohort from SimData
set.seed(12321)
#create a sample index. 1 if sampled, 0 if not
N <- nrow(SimData)
sampleInd <- rep(0, N)

# sample all with observed failure time. (200 individuals)
sampleInd[SimData$status==1] <- 1

#sample 150 more observations from the entire data set without replacement
sampleInd[sample(1:N, 150)] <- 1

table(sampleInd) #total number of subcohort is 293 
```

Now we calculate the sample weights $w_i = 1/Pr(Sampled from cohort)$ for each observation included in the sub-cohort. 
```{r}

#first calculate the Pr(Sampled from cohort) for each observation
sampleProb <- numeric(500)

#all non-censored observations were sampled, so their sample probability is 1
sampleProb[SimData$status==1] <- 1 
sampleProb[SimData$status==0] <- 150/N

SimData$weights <- 1/sampleProb

subCohortData <- SimData[sampleInd==1,]
```

Here we estimate all measures available. 

```{r}
#estimate accuracy measures using only the subcohort data
survMTP.estimate(time =subCohortData$survTime, 
                event = subCohortData$status, 
                marker = subCohortData$Y,
                weights = subCohortData$weights,
                 cohortN = N,
                study.design = "Case-Cohort",
                predict.time = 2, 
                 cutpoint = 0)

```

Only estimate the $AUC$ and $TPR(0)$. 

```{r}
tmp <- survMTP.estimate(time =subCohortData$survTime, 
                event = subCohortData$status, 
                marker = subCohortData$Y,
                weights = subCohortData$weights,
                cohortN = N,
                study.design = "Case-Cohort",
                predict.time = 2, 
                measures = c("AUC", "TPR"),
                cutpoint = 0)
tmp
```

```{r}
#access the estimates 
tmp$estimates

#and the confidence bounds
tmp$CIbounds
```

####Type II: 
 
In the second sampling design, we sample $n=100$ from the strata of participants with observed failures and $n=100$ from censored individuals. We need to account for this when we calculate the sampling weights. 

```{r}

#create a new sample index. 1 if sampled, 0 if not
sampleInd <- rep(0, N)

# sample 100 individuals with observed failure time. 
CaseInd <- c(1:N)[SimData$status==1] 
sampleInd[sample(CaseInd, 100)] <- 1

# sample 100 individuals with censored failure time. 
ControlInd <- c(1:N)[SimData$status==0] 
sampleInd[sample(ControlInd, 100)] <- 1

table(sampleInd) #total number of subcohort is 200 
```

Calculate the sample weights.

```{r}

#first calculate the Pr(Sampled from cohort) for each observation
sampleProb <- numeric(500)

#all non-censored observations were sampled
sampleProb[SimData$status==1] <- 100/length(CaseInd) 
sampleProb[SimData$status==0] <- 100/length(ControlInd)

SimData$weights2 <- 1/sampleProb

subCohortData2 <- SimData[sampleInd==1,]
```

Estimate all measures available. 

```{r}
#estimate accuracy measures using only the subcohort data
survMTP.estimate(time =subCohortData2$survTime, 
                event = subCohortData2$status, 
                marker = subCohortData2$Y,
                weights = subCohortData2$weights2,
                cohortN = N,
                study.design = "Case-Cohort",
                predict.time = 2, 
                 cutpoint = 0)

```
For more information see `?survMTP.estimate`. 


### References
Liu D, Cai T, Zheng Y. Evaluating the predictive value of biomarkers with stratified case-cohort design. *Biometrics* 2012, 4: 1219-1227.

Pepe MS, Zheng Y, Jin Y. Evaluating the ROC performance of markers for future events. *Lifetime Data Analysis.* 2008, 14: 86-113.

Zheng Y, Cai T, Pepe MS, Levy, W. Time-dependent predictive values of prognostic biomarkers with failure time outcome. *JASA* 2008, 103: 362-368.














